---
title: "The envnames package helps navigate user-defined and function execution environments, and find objects in nested environments"
author: "Daniel Mastropietro (mastropi@uwalumni.com)"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Overview of the envnames package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
#library(knitr)
#opts_chunk$set(include=TRUE, warning=FALSE)
```


```{r echo=FALSE}
library(envnames)
rm(list=ls())
```

# Introduction

The main goal of this package is to overcome the limitation of the `environmentName()` function in the base package which does not return the name of an environment unless it is a package, a namespace or a pre-defined environment (e.g. the global environment, the base environment). In fact, the `environmentName()` function returns an *empty string* when the argument is a user-defined environment.

On the other hand, the environment itself is identified solely by its memory address, which makes it difficult to track an environment once we have defined a number of them. These limitations --and the workaround provided by this package-- can be seen by running the following code snippet:  

```{r Motivation}
env1 <- new.env()
cat("The name of the environment just defined is: ", environmentName(env1), "(empty)\n")
cat("Simply referencing the environment just defined yields its memory address,
    which is not so helpful:"); print(env1)
cat("Using the environment_name() function of the envnames package gives
    the environment name:", environment_name(env1))
```

Clearly the last one is the result we most likely want, and the envnames package makes this possible by creating a lookup table that maps environment names to their memory addresses. Using this lookup table, it is possible to retrieve the name of any environment where an object resides, _even_ the name of function execution environments (in which case, the name is given by the function itself).

Why do we care about knowing the name of user-defined environments and function execution environments? That piece of information may be handy for example under the following scenarios:

- working in a package where user-defined environments have been defined in a nested structure:  
this package facilitates the navigation through those environments and their connection between them, eliminating e.g. the use of `ls()` as a rudimentary tool to identify the human-understandable environment (e.g. `env1`) referred by an environment given by its memory address (e.g. `<environment: 0x063dbc90>` in 32-bit systems or `<environment: 0x00000000063dbc90>` in 64-bit systems), as already seen above.  
For more information and examples, see below the section on function `get_env_names()`, that returns a map of currently defined environments and the way they are connected or nested.  
- debugging an application:  
this package makes it easier to retrieve variables in different environments; for instance, retrieve the value of a variable in the parent environment to the environment where the debugger is currently positioned, which could well be a function execution environment.  
For more information and examples, see below the sections on functions `get_obj_name()` and `get_obj_value()`, which can be used to retrieve the name and value of the variable leading to a particular variable in connected environments, and the section on `environment_name()`, which can be used to retrieve the name of an execution environment.  

Apart from this core functionality, additional tools were added during the package development process which include:

- an enhancement of the built-in `exists()` function with the capability of searching objects *recursively* --i.e. in  environments defined inside *other* environments--, as well as searching objects that are the *result of expressions*. This functionality is provided by the `obj_find()` function.  
- a simplification of the output obtained when retrieving the calling function name and the stack of calling functions, currently provided by the built-in function `sys.call()`. This functionality is provided by functions `get_fun_name()`, `get_fun_calling()`, and `get_fun_calling_chain()`, which return simple strings or array of strings with the function names of interest.  
- the retrieval of the memory address of an object. This functionality is provided by the `get_obj_address()` function.  

Currently the package has 11 functions directly accessible to the user (plus one function that is an alias).

*Naming convention*: Function names are all small caps and the underscore is used to separate keywords (e.g. `environment_name()`, `get_obj_address()`, etc.)

# Description of the functions

This section describes the functionality of the 11 available functions (sorted by relevance based on a mix of historical and practical use view):  

1) `get_env_names()`, used to retrieve the name of all the environments defined in the workspace together with their memory address. This is an address-name lookup table, the core element of the package that allows the "magic" to happen.  

2) `environment_name()` / `get_env_name()` (its alias), used to get the name of user-defined and execution environments (as well as all other named environments).  

3) `obj_find()`, used to find an object in the workspace and recursively within environments.  

4) `get_fun_name()`, `get_fun_calling()`, `get_fun_calling_chain()`, used to get the function calling name and stack displayed in a format that is easier than the one provided by `sys.calls()`.  

5) `get_fun_env()`, used to retrieve a function's execution environment.  

6) `get_obj_name()`, `get_obj_value()`, used to retrieve the *name* and *value* of the object leading to a given function's parameter.  

7) `get_obj_address()`, used to get the memory address of an object by first looking for it (using `obj_find()`).  

8) `address()`, used to return the memory address of an object existing in the environment where the function is run.  

In the following sub-sections, each title is a sentence stating the main purpose of the presented function.

## 1) `get_env_names()`: retrieve the address-name lookup table of defined environments
The `get_env_names()` function returns the address-name lookup table that is the basis for most of the other functions in the package, which rely on it to retrieve the names of environments based on their memory addresses.

The signature of the function is the following:  
`get_env_names(envir = NULL, include_functions = FALSE)`.

### Examples

#### Let's start with the definition of a few environments
We define a couple of environments and nested environments.

```{r DefineEnvironments}
env1 <- new.env()
env_of_envs <- new.env()
with(env_of_envs, env21 <- new.env())
```

Note that environment `env21` is *nested* in environment `env_of_envs`.

#### Basic operation
The following call returns a data frame containing the address-name lookup table, where the two main columns are `address` (containing the memory address of the environment) and `name` (containing the name of the environment). The other columns are used to give context to the environments, such as `path` which tells us how to get to nested user-defined environments (see the case for environment `env21` nested within `env_of_envs`), or their `location`, which for instance indicates the package where an environment is defined, or the environment where a function is defined (if the concerned environment is a function's execution environment).

```{r GetLookupTable, warning=FALSE}
get_env_names()
```

We can also restrict the lookup table to the environments (including function execution environments) defined within another environment. Now the `path` to the environment is relative to the environment on which the search is restricted.

```{r GetLookupTableRestricted, warning=FALSE}
get_env_names(envir=env_of_envs)
```

## 2) `environment_name()`: retrieve name of user-defined and function execution environments
The `environment_name()` function (or its alias `get_env_name()`) extends the functionality of the built-in `environmentName()` function by also retrieving the name of user-defined environments and function execution environments.
Although the name of an environment can be easily retrieved with `deparse(substitute(env1))` where `env1` is a user-defined environment, the most useful scenario is when we have just the *memory address* of the environment where e.g. an object resides (as in e.g. `<environment: 0x0437fb40>` in 32-bit systems or `<environment: 0x000000000437fb40>` in 64-bit systems). In this scenario, `environment_name()` can tell us the *name* of the
environment having that memory address.  

Note that the address-to-name resolution also works for *function execution environments*, as we shall see in the examples below.

The signature of the function is the following:  
`environment_name(env, envir = NULL, envmap = NULL, matchname = FALSE, ignore = NULL, include_functions = FALSE)`.

### Examples

#### Basic operation
Let's retrieve the names of the environments defined above. This may sound trivial because we are already typing the environment name! However, look at the output from the second call: it contains the *path* to the environment being searched for stating that `env21` is found inside environment `env_of_envs`.

```{r GetEnvironmentNames, warning=FALSE}
environment_name(env1)
environment_name(env21)
```

If we already know in which environment the environment we are after is defined, we can specify it in the `envir` parameter and get their name without any path information:

```{r GetEnvironmentNamesSpecifyingLocation, warning=FALSE}
environment_name(env1, envir=globalenv())
environment_name(env21, envir=env_of_envs)
```

#### More advanced examples
Suppose we now define another environment that points to one of the already defined environments.

```{r DefineAnotherEnvironmentThatPointsToAnExistingEnvironment}
e_proxy <- env_of_envs$env21
```

Now, let's retrieve its name:

```{r GetTheEnvironmentNameOfThisNewEnvironment}
environment_name(e_proxy)
```

What we get is a named array containing names of *all* the environments (in alphabetical order) that point to the same memory address (with their location as the array names). We can disable such behaviour of matching environments by memory address by setting the `matchname` parameter to `TRUE` so that the returned environments must match both the memory address *and* the given name:

```{r CallEnvironmentNameWithMatchNameTRUE1}
environment_name(e_proxy, matchname=TRUE)
```

Now the result is an *unnamed* array because there is only one environment matched by the search for the `e_proxy` environemnt. Furthermore, the result indicates that the environment is defined in the global environment, as otherwise the location where it were defined would be part of the name (as in e.g. `env1$e_proxy`).  

Note however that the last call could actually return *more than one environment* in the unlikely scenario where environments sharing the same name (`e_proxy` in the above example) are defined in different environments *all pointing to the same environment (memory address)*. In the above example we could have this situation if we defined an environment called `"e_proxy"` in environment `env_of_envs` pointing to environment `e_proxy` defined in the global environment.
Such scenario would yield:

```{r CallEnvironmentNameWithMatchNameTRUE2}
env_of_envs$e_proxy <- e_proxy
environment_name(e_proxy, matchname=TRUE)
```
i.e., again a named array with all the matches to the searched environment.

Finally, if we try to retrieve the environment name of a non-existing environment, we get `NULL`.

```{r CallEnvironmentNameOnNonExistingEnvironment, warning=FALSE}
environment_name(non_existing_env)
```

#### Retrieving the environment name associated with a memory address
Now suppose we have a memory address and we would like to know if that memory address represents an environment.
We can simply call `environment_name()` with the memory address as argument, as shown in the following example:

```{r ConvertMemoryAddressToEnvironmentName}
env1_address = get_obj_address(env1)
environment_name(env1_address)
```

Of course, in practice we would not call the `get_obj_address()` function to get the environment's memory address; we would simply type in the memory address we are after. Note that this memory address depends on the architecture (32-bit or 64-bit) and it can be given in one of the following four ways:  

- an 8-digit (32-bit) / 16-digit (64-bit) address, e.g. `"0000000011D7A150"` (64-bit architecture)  
- a 10-digit (32-bit) / 18-digit (64-bit) address, e.g. `"0x0000000011D7A150"` (64-bit architecture)  
- either of the above addresses enclosed in `< >`, e.g. `"<0000000011D7A150>"` or `"<0x0000000011D7A150>"` (64-bit archiecture)  
- a 10-digit (32-bit) / 18-digit (64-bit) address preceeded by the `environment: ` keyword and enclosed in `< >`, e.g.: `"<environment: 0x0000000011D7A150>"` (64-bit architecture)  

(note: Linux Debian distributions may have a 12-digit memory address representation. The best way to know
what the memory address representation is in a particular system is to call e.g. `address("x")`.)

The last format is particularly useful when copying & pasting the result of querying an environment object, for example when typing `env1` at the `R` command prompt, in which case we get:

```{r}
env1
```

If the memory address does not match any of the above formats or does not represent an environment, `environment_name()` returns `NULL`. Ex:

```{r EnvironmentNameOfNonEnvironmentMemoryAddressIsNULL}
x = 2
environment_name(get_obj_address(x))
```

as the address of `x` is not the address of an environment.

#### Retrieving a function execution environment
If called from within a function with no arguments, `environment_name()` returns the execution environment of the function, which is identified by the name of the function. This is given with its *full path*, as in e.g. `env1$f`, when `environment_name()` is called from function `f()` defined in environment `env1`.

Since the first argument of `environment_name()` is the environment whose name we want to retrieve, we could also retrieve the execution environment of any calling function by specifying the corresponding `parent.frame`. Once again the name of such parent execution environment would be the name of the function given with its *full path*. 

The following example illustrates the above two use cases.

```{r GetExecutionEnvironmentName1}
with(env_of_envs$env21, {
  f <- function() {
    cat("1) We are inside function:", environment_name(), "\n")
    cat("2) The calling environment is:", environment_name(parent.frame()), "\n")
  }
  g <- function() {
    f()
  }
})
cat("Having defined both f() and g() in environment env_of_envs$env21,
    and having function g() call f()...\n")
cat("...when we call env_of_envs$env21$f() from the global environment,
    we get the output that follows:\n")
env_of_envs$env21$f()
cat("\n...and when we call f() from inside function g(),
    we get the output that follows:\n")
env_of_envs$env21$g()
```
Note that, in the second case, the enviroment showing as path to `f()` is *not* `env_of_envs$env21` (as we would have expected) but `e_proxy`. The reason is that environment `e_proxy` (in the global environment) points to the same memory address as `env_of_envs$env21`. And since environment names are retrieved by their memory address  (which in this case is the memory address of `f`'s execution environment), there may be more than one environment matching the same memory address. In such cases, the rule implemented in `environment_name()` is to retrieve the matching environment whose name comes first in alphabetical order (which in this case is `e_proxy` --coming before both `env_of_envs$e_proxy` and `env_of_envs$env21` in alphabetical order, all environments that match the memory address of the environment where `f()` is defined).

But if we call `env_of_envs$env21$f()` (instead of calling `f()` as above) from a function `h()` defined in the `env_of_envs$env21` environment, we get:


```{r GetExecutionEnvironmentName2}
with(env_of_envs$env21, {
  f <- function() {
    cat("1) We are inside function", environment_name(), "\n")
    cat("2) The calling environment is:", environment_name(parent.frame()), "\n")
  }
  h <- function() {
    env_of_envs$env21$f()
  }
  }
)
env_of_envs$env21$h()
```

## 3) `obj_find()`: check the existence of (visible) objects
With the `obj_find()` function we can check if an object exists in any existing environment (as long as it has been exported in the case of package environments). All environments --including system environments, packages, user-defined environments, and optionally function execution environments-- are crawled and searched for the object. This includes any environments that are defined _within_ other environments (_nested_), and it represents an enhancement to the built-in `exists()` function. The function returns a character array with all the environments where the object has been found.

Objects to search for can be specified either as a symbol or as a string. Ex: `obj_find(x)` and `obj_find("x")` both look for an object called "x". They can also be the result of an expression as in `v[1]`.

The function returns `NULL` if the object is not found or if the expression is invalid. For instance `obj_find(unquote(quote(x)))` returns `NULL` because the `unquote()` function does not exist in R.

The signature of the function is the following:  
`obj_find(obj, envir = NULL, envmap = NULL, globalsearch = TRUE, n = 0, return_address = FALSE, include_functions = FALSE, silent = TRUE)`

### Examples

#### Let's start with a few object definitions
We define a couple of objects in the environments previously defined:

```{r DefineObjectsInEnvironments}
x <- 5
env1$x <- 3
with(env_of_envs, env21$y <- 5)
with(env1, {
  vars_as_string <- c("x", "y", "z")
})
```

#### Basic operation
Now let's look for these objects:
```{r LookForObjects1, warning=FALSE}
environments_where_obj_x_is_found = obj_find(x)
cat("Object 'x' found
in the following environments:"); print(environments_where_obj_x_is_found)
environments_where_obj_y_is_found = obj_find(y)
cat("Object 'y' found
in the following environments:"); print(environments_where_obj_y_is_found)
```

(if we are seeing more environments than expected in the above output, let us recall that two `e_proxy` environments point to the same environment as `env_of_envs$env21`)

```{r LookForObjects2}
environments_where_obj_is_found = obj_find(vars_as_string)
cat("Object 'vars_as_string' found
in the following environments:"); print(environments_where_obj_is_found)
```

Let's also look for the objects defined in `vars_as_string` and `vars_quoted`.
```{r LookForObjectsWhoseNamesAreGivenInArray}
environments_where_obj_1_is_found = obj_find(env1$vars_as_string[1])
  ## Here we are looking for the object 'x'
cat(paste("Object '", env1$vars_as_string[1], "' found
in the following environments:")); print(environments_where_obj_1_is_found)
environments_where_obj_2_is_found = obj_find(env1$vars_as_string[2])
  ## Here we are looking for the object 'y'
cat(paste("Object '", env1$vars_as_string[2], "' found
in the following environments:")); print(environments_where_obj_2_is_found)
environments_where_obj_3_is_found = obj_find(env1$vars_as_string[3])
  ## Here we are looking for the object 'z' which does not exist
cat(paste("Object '", env1$vars_as_string[3], "' found
in the following environments:")); print(environments_where_obj_3_is_found)
```

or using `sapply()` to look for all the objects whose names are stored in `env1$vars_as_strings` at once:

```{r LookForObjectsUsingSAPPLY}
environments_where_objs_are_found = with(env1, sapply(vars_as_string, obj_find) )
cat("The objects defined in the 'env1$vars_as_string' array are found
    in the following environments:\n");
print(environments_where_objs_are_found)
```

Note how calling `obj_find()` from within the `env1` environment (which we do in order to resolve the `vars_as_string` variable --the argument of `obj_find()`) still searches for the objects everywhere. This is because parameter `globalsearch` is set to `TRUE`. If we set it to `FALSE` and we add `envir=env1` as searching environment, we would get:  

```{r LookForObjectsUsingSAPPLYNoGlobalSearch}
environments_where_objs_are_found = with(env1,
                      sapply(vars_as_string, obj_find, globalsearch=FALSE, envir=env1) )
cat("The objects defined in the 'env1$vars_as_string' array are found
    in the following environments (no globalsearch):\n");
print(environments_where_objs_are_found)
```
NOTE: Even if we are running `sapply()` inside environment `env1`, it's important to add parameter `envir=env1` to the call to `obj_find()`; if we don't add it, the objects are *not* found because the parent environment of the call to `obj_find()` is *not* `env1` but the `sapply()` execution environment, where the objects do not exist.

We can also search for objects given as a symbol:

```{r LookForObjectAsASymbol}
environments_where_obj_x_is_found = obj_find(as.name("x"))
cat("Object 'x' found
in the following environments:"); print(environments_where_obj_x_is_found)
```

Finally, we can also search for visible (exported) objects defined in packages:

```{r LookForObjectsDefinedInPackages}
environments_where_obj_is_found = obj_find(aov)
cat("Object 'aov' found
in the following environments:"); print(environments_where_obj_is_found)
```

## 4) `get_fun_name()`, `get_fun_calling()`, `get_fun_calling_chain()`: retrieve functions in the function calling chain (stack)
Functions `get_fun_name()`, `get_fun_calling()`, and `get_fun_calling_chain()` can be used to retrieve information about
calling functions. The first two retrieve information about one function while the latter retrieves information about the functions in the calling chain or stack.

Following is an example that shows the difference between using the package function `get_fun_calling_chain()` vs. the built-in function `sys.call()` to retrieve the calling stack. 

In particular note:  
- How easy it is to check what the calling function is (just do a string comparison as in e.g. `get_fun_calling() == "env1$f"`). On the contrary, when using `sys.call()` we first need to parse the output before making such a comparison. See **[this link](http://stackoverflow.com/questions/15595478/how-to-get-the-name-of-the-calling-function-inside-the-called-routine)** for more details.  
- We get a data frame containing the chain of calling functions, from the most recent call to least recent, including function parameters if desired.  

The signatures of the three aforementioned functions are:  
`get_fun_name(n = 0)`  
`get_fun_calling(n = 1, showParameters = FALSE)`  
`get_fun_calling_chain(n = NULL, showParameters = FALSE, silent = TRUE)`

### Examples

#### Let's start with a few object definitions
1) First we define a couple of new environments:

```{r DefineTwoEnvironments}
env11 <- new.env()
env12 <- new.env()
```

2) Now we define an example function `h` to be called by two different functions `f` defined in two different user-environments. This function `h` sums `+1` or `+2` to the input parameter `x` depending on which function `f` was responsible for calling it.

```{r DefineFunctionH}
with(globalenv(), 
h <- function(x, silent=TRUE) {
  fun_calling_chain = get_fun_calling_chain(silent=silent)

  # Do a different operation on input parameter x depending on the calling function
  fun_calling = get_fun_calling(showParameters=FALSE)
  if (fun_calling == "env11$f") { x = x + 1 }
  else if (fun_calling == "env12$f") { x = x + 2 }

  return(x)
}
)
```

3) Finally we define the two functions `f` that call `h`, respectively in environments `env11` and `env12`:

```{r DefineTwoFunctionsFInSeparateEnvironments}
with(env11,
  f <- function(x, silent=TRUE) {
    fun_calling_chain = get_fun_calling_chain()
    return(h(x, silent=silent))
  }
)

with(env12,
  f <- function(x, silent=TRUE) {
    fun_calling_chain = get_fun_calling_chain()
    return(h(x, silent=silent))
  }
)
```

### Basic operation
We now run these functions `f` and take note of their output.

- Output from `env11$f()`:

```{r RunFunctionF1, echo=FALSE}
silent = FALSE
x = 0
cat("\nWhen h(x) is called by env11$f(x=", x, ") the output is: ", env11$f(x, silent=silent), "\n", sep="")
```

- Output from `env12$f()`:

```{r RunFunctionF2, echo=FALSE}
silent = FALSE
x = 0
cat("\nWhen h(x) is called by env12$f(x=", x, ") the output is: ", env12$f(x, silent=silent), "\n", sep="")
```

Note how easy it was (by using just a string comparison) to decide what action to take based on the `f()` function calling `h()` and perform a different operation.

Note also that, in order to decide between the two possible calling functions `env11$f()` or `env12$f()` we used `get_fun_calling()`, as opposed to `get_fun_name()`, because the latter returns *just* the function name, devoided of any environment name.

## 5) `get_fun_env()`: retrieve a function's execution environment
The `get_fun_env()` function can be used to easily place oneself in a function's execution environment and explore it.

The following example illustrates.

### Basic operation
Let's start defining a couple of functions that make up a function calling chain. The called function `h()` retrieves the memory address of its callee.

```{r GetFunEnv}
h <- function(x) {
  # Get the value of parameter 'x' in the execution environment of function 'env1$g'
  # The returned value is a list because there may exist different instances of the same function.
  xval_h = x
  xval_g = evalq(x, get_fun_env("env1$g")[[1]])
  cat("The value of variable 'x' in function", get_fun_name(), "is", xval_h, "\n")
  cat("The value of variable 'x' inside function env1$g is", xval_g, "\n") 
}
env1 <- new.env()
with(env1, 
  g <- function() {
    x = 2
    return( h(3) )
  }
)
env1$g() 
```
When `get_fun_env()` is called from outside a function, it returns NULL, even when the function exists.

```{r GetFunEnvOutside}
cat("The execution environment of a function that is not in the calling chain is:\n")
print(get_fun_env("env1$g"))

```

### Advanced example that puts together `get_fun_name()` and `get_fun_env()`
```{r GetFunEnvCombinedExample}
h <- function(x) {
  # Get the value of parameter 'x' in the execution environment of function 'env1$g'
  # The returned value is a list because there may exist different instances of the same function.
  parent_function_name = get_fun_name(n=1)
  cat("Using get_fun_name() and environment_name() functions:
      The parent frame of function", get_fun_name(), "is", get_fun_name(n=1), "\n")
  xval_h = x
  xval_g = evalq(x, get_fun_env(parent_function_name)[[1]])
  cat("Using get_fun_name():
      The value of variable 'x' in function", get_fun_name(), "is", xval_h, "\n")
  cat("Using get_fun_env() and evalq() functions:
      The value of variable 'x' inside function", parent_function_name, "is", xval_g, "\n") 
}
env1 <- new.env()
with(env1, 
  g <- function() {
    x = 2
    return( h(3) )
  }
)
env1$g() 
```

## 6) `get_obj_name()`, `get_obj_value()`: retrieve the name/value of an object at a specified parent generation
The `get_obj_name()` and `get_obj_value()` functions are intended to help track objects and their values as they are passed through different environments. The most useful of the two is `get_obj_name()`, because the *values* of linked objects are the same as they traverse the different environments, making `get_obj_value()` be almost the same as calling `deparse()` at any environment (except for some special cases described in the documentation).

When called from within a function `get_obj_name()` can be used to know the name of the object that *leads* to a particular parameter a few generations back following the function calling chain. In other words, it helps us know the object in a given parent generation that is "responsible" for a function's parameter value.

After learning a little more about `get_obj_name()`, one may have the impression that it gives the same result as the one provided by `deparse(substitute())`. However, this is not the case as is shown in the examples that follow.

The signatures of these two functions are:  
`get_obj_name(obj, n = 0, eval = FALSE, silent = TRUE)`
`get_obj_value(obj, n = 0, silent = TRUE)`

### Examples

#### Let's start with a few function definitions
```{r GetObjNameExampleDefinitions}
getObjNameAndCompareWithSubstitute <- function(y, eval=FALSE) {
  parent_generation = 2
  get_obj_name_result = get_obj_name(y, n=parent_generation, eval=eval)
  deparse_result = deparse(y)
  substitute_result = substitute(y, parent.frame(n=parent_generation))
  deparse_substitute_result = deparse(substitute(y, parent.frame(n=parent_generation)))
  eval_result = evalq(y, envir=parent.frame(n=parent_generation))
  if (!eval) {
    cat("Result of get_obj_name(y, n=", parent_generation, "): ", get_obj_name_result,
        "\n\tConceptually this is the name of the object at parent generation ",
        parent_generation,
        "\n\tLEADING to *parameter* 'y'.\n", sep="")
    cat("Result of deparse(substitute(y, parent.frame(n=", parent_generation, "))): ",
        deparse_substitute_result,
        "\n\tConceptually this is the substitution of *variable* 'y'
        at parent generation ", parent_generation,
        "\n\tconverted to a string.\n", sep="")
  } else {
    cat("Result of get_obj_name(y, n=", parent_generation, ", eval=", eval, "): ",
        get_obj_name_result,
        "\n\tConceptually this is the object LEADING to *parameter* 'y' evaluated
        at parent generation ", parent_generation, ".\n", sep="")
    cat("Result of deparse(y): ", deparse_result,
        "\n\tConceptually this is the value of *parameter* 'y' converted to a character
        string.\n", sep="")
    cat("Result of substitute(y, parent.frame(n=", parent_generation, ")): ",
        substitute_result,
        "\n\tConceptually this is the substitution of *variable* 'y' at parent generation ",
        parent_generation,
        ".\n", sep="")
    cat("Result of evalq(y, envir=parent.frame(n=", parent_generation, ")): ",
        eval_result,
        "\n\tConceptually this is the evaluation of *variable* 'y' at parent generation ",
        parent_generation,
        ".\n", sep="")
  }
}

callGetObjNameAndCompareWithSubstitute <- function(x, eval=FALSE) {
  getObjNameAndCompareWithSubstitute(x, eval=eval)
}
```

#### Basic operation
Let's compare the result of calling `get_obj_name()` with the result of `deparse(substitute())`:

```{r GetObjNameExampleCall1}
y <- -9   # Global variable with the same name as the parameter of testing function
z <- 3
callGetObjNameAndCompareWithSubstitute(z)
```

Note the conceptual difference: `deparse(substitute(y, parent.frame(n=2)))` retrieves the object assigned to `y` at parent generation 2 (`substitut`ion) and returns it as a string (`depars`ing), while `get_obj_name(y, n=2)` **first** *traces back* the object names in parent generations leading to parameter `y`, and **then** returns the name of the object at the specified parent generation.

When `eval=TRUE`, `get_obj_name()` behaves the same way as `deparse()`, because the values of the objects leading to parameter `y` in parent generations is always the same and equal to the parameter's value. This result is the same as the one obtained by calling `get_obj_value()`. The following example illustrates:
```{r GetObjNameExampleCall2}
y <- -9   # Global variable with the same name as the parameter of testing function
z <- 3
callGetObjNameAndCompareWithSubstitute(z, eval=TRUE)
```
That is, calling `get_obj_name(y, n=n, eval=TRUE)` (or its equivalent `get_obj_value()`) retrieves the value of parameter `y` in parent generation `n`, which is the same in all parent generations and equal to the value of parameter `y` inside the calling function. Therefore, this is the same as the result of `deparse(y)`.
On the other hand substituting or evaluating variable `y` in parent generation 2 concerns directly variable `y` in *that* parent generation.

#### Finding the parameter path leading to a given function's parameter
The `get_obj_name()` function can also be used to find the set of variables in the different parent environments leading to a specified variable in the current environment. A particular case of this is the parameter path in a function calling chain leading to a function's parameter, which is illustrated below.

Let's define a set of simple functions that create a calling chain, `f1() -> f2() -> f3()` each of them having a parameter with a different names:

```{r RetrieveParameterPath}
f1 <- function(x) {
  cat("f1(x) is calling f2(y=x)...\n")
  f2(x)
}
f2 <- function(y) {
  cat("f2(y) is calling f3(z=y)...\n")
  f3(y)
}
f3 <- function(z) {
  cat("f3(z) is retrieving the parameter path from three parent environments
      leading to function parameter z...\n\n")
  cat("Output from get_obj_name(z, n=3, silent=FALSE):\n")
  variable_leading_to_z_3levels_back = get_obj_name(z, n=3, silent=FALSE)
}
w = 1.3
f1(w)
```
So, we clearly see the environments and variables leading to parameter `z` from `R_GlobalEnv$w`:  
`R_GlobalEnv$w -> f1$x -> f2$y -> f3$z`

### Use of `get_obj_value()`
The result of calling `get_obj_value()` is the same as calling `get_obj_name()` with `eval=TRUE`.
It may come as a handy function (by reducing writing) to use in debugger contexts to find out the value of variables in different environments.

The following example illustrates the use of the function from within a function and shows the difference with the result of using `evalq()`.
Let's start defining two functions:

```{r GetObjValueExampleDefinitions}
getObjValueAndCompareWithEval <- function(y) {
  parent_generation = 2
  get_obj_value_result = get_obj_value(y, n=parent_generation)
  eval_result = evalq(y, envir=parent.frame(n=parent_generation))
  cat("Result of get_obj_value(y, n=", parent_generation, "): ", get_obj_value_result,
      "\n\tConceptually this is the object LEADING to *parameter* 'y' 
      evaluated at parent generation ",
      parent_generation, ".\n", sep="")
  cat("Result of evalq(y, envir=parent.frame(n=", parent_generation, ")): ", eval_result,
        "\n\tConceptually this is the evaluation of *variable* 'y' at parent generation ",
        parent_generation, ".\n", sep="")
}

callGetObjValueAndCompareWithEval <- function(x) { getObjValueAndCompareWithEval(x) }
```

Now let's see the results of calling this function which explains the differences between `get_obj_value()` and `evalq()`.

```{r GetObjValueExampleCall}
y <- -9   # Global variable with the same name as the parameter of testing function
z <- 3
callGetObjValueAndCompareWithEval(z)
```

## 7 & 8) `get_obj_address()` and `address()`: retrieve the memory address of an object
Following are examples of using the `get_obj_address()` function to retrieve the memory address of an object, which is then checked by the `address()` function that calls the direct method (via a C function call) to retrieve an object's memory address. The differences between these two functions are also explained.

In the `get_obj_address()` function, the object can be given either as a symbol or as an expression. If given as an expression, the memory address of the _result_ of the expression is returned. If the result is yet _another_ expression, the process stops, i.e. the memory address of that final expression is returned.

Internally this funcion first calls `obj_find()` to look for the object (using `globalsearch=TRUE`) and then retrieves the object's memory address, showing the name of all the environments where the object was found, or `NULL` if the object is not found.  

The signature of the function is the following:  
`get_obj_address(obj, envir = NULL, envmap = NULL, n = 0, include_functions = FALSE)`

### Examples
The following two calls return the same result:

```{r GetObjectAddress1}
obj_address1 = get_obj_address(x)
cat("Output of 'get_obj_address(x)':\n"); print(obj_address1)
obj_address2 = with(env1, get_obj_address(x))
cat("Output of 'with(env1, get_obj_address(x))':\n"); print(obj_address2)
```

Note especially the last case, where calling `get_obj_address()` from within the `env1` environment still searches for the object everywhere.

We can restrict the memory address returned by making the environment where the object is located explicit --by either using the `$` notation or the `envir` parameter of `get_obj_address()`. In this case only the address of the specified object is returned, even if other objects with the same name exist within the specified environment. A few examples follow:

```{r GetObjectAddress2}
get_obj_address(env1$x)
get_obj_address(x, envir=env1)
with(env1, get_obj_address(x, envir=env1))
```

Note there is a slight difference between calling `get_obj_address()` using the `$` notation and calling it with the `envir=` parameter: in the latter case, the result is an _unnamed_ array.    

&nbsp;
&nbsp;

Suppose now the object is an expression referencing three potential existing objects as strings, more specifically an array:

```{r GetNonExistentObjectAddress}
vars = c("x", "y", "nonexistent")
get_obj_address(vars[1], envir=env1)
sapply(vars, get_obj_address)
```

(if we are seeing more environments than expected in the above output, let us recall that environment `e_proxy` points to the same environment as `env_of_envs$env21`)

We can check that the memory address is correct by running the internal function `address()` which calls a C function that retrieves the memory address of an object:

```{r CheckMemoryAddressIsCorrect}
address(env1$x)
address(e_proxy$y)
```

Finally: why would we use `get_obj_address()` instead of `address()` to retrieve the memory address of an object? For two main reasons:  
- `get_obj_address()` first searches for the object in all user-defined environments, while `address()` needs to be called from within the environment where the object is defined.  
- `get_obj_address()` returns `NULL` if the object does not exist, while `address()` returns the _memory address_ of the `NULL` object, which may be misleading.  

To prove the second statement, we simply run the following two commands which yield the same result:

```{r MemoryAddressOfNULL}
address(env1$nonexistent)
address(NULL)
```
while running `get_obj_address()` on the non-existent object yields `NULL`:

```{r MemoryAddressOfNonExistentIsNULL}
get_obj_address(env1$nonexistent)
```

# Summing up
We have described all the 11 visible functions defined in the `envnames` package and shown examples of using them:  

1) `get_env_names()` to retrieve all the environments defined in the workspace in the form of a lookup table where the environment name can be looked up from its memory address.  

2) `environment_name()` / `get_env_name()` (its alias) to retrieve the name of an environment. This function extends the functionality of the built-in `environmentName()` function by retrieving:  
    - the name of a user-defined environment  
    - the name and path to environments defined inside other environments  
    - the name and path to the function associated to an execution environment  
    - the name of the environment associated to a memory address  

3) `obj_find()` to find an object in the workspace. This function extends the functionality of the built-in `exists()` function by:  
    - searching for the object in user-defined environments  
    - searching for the object recursively (i.e. in environments defined inside other environments)  

4) `get_fun_name()`, `get_fun_calling()`, `get_fun_calling_chain()` to get the stack of calling functions. These functions return the stack information in a manner that is much simpler than the built-in `sys.calls()` function, making it easier to check the _names_ of the calling functions.  

5) `get_fun_env()` to get the execution environment of a given function in the calling stack, so that we can query objects that live within.  

6) `get_obj_name()`, `get_obj_value()` to retrieve the name and value of the object leading to a given function's parameter.

7) `get_obj_address()` to retrieve the memory address of an object. This function gives a functionality that is not available in base R. Note that the `data.table` package also provides a function called `address()` to retrieve the memory address of an object; however the object is _not searched for in the whole workspace_ as is the case with the `get_obj_address()` in this package.  

8) `address()` to retrieve the memory address of an existing object, but without looking for it first.  

&nbsp;
&nbsp;

*This vignette was generated under the following platform:  *
```{r SystemInfo, echo=FALSE}
data.frame(SystemInfo=Sys.info()[c("sysname", "release", "version", "machine")])
version
```
