---
title: "Overview of the envnames package"
author: "Daniel Mastropietro"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Overview of the envnames package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
#library(knitr)
#opts_chunk$set(include=TRUE, warning=FALSE)
```


```{r echo=FALSE}
library(envnames)
```

The main goal of this package is to overcome the limitation of the built-in `environmentName()` function from the base package which does not return the name of an environment unless it is a package, a namespace or a pre-defined environment (e.g. the global environment, the base environment). The `environmentName()` function returns an empty string for user-defined environments.

As mentioned in the **[documentation for the `environment()` function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/environment.html)**:  
*System environments, such as the base, global and empty environments, have names as do the package and namespace environments and those generated by attach(). Other environments can be named by giving a "name" attribute, but this needs to be done with care as environments have unusual copying semantics.*

The envnames package solves this problem by creating a lookup table that maps environment names to their memory addresses. Using this lookup table, it is possible to retrieve the name of any environment where an object resides, including function execution environments.

Apart from this core functionaliy, additional tools were added during the package development process which include:

- an enhancement of the built-in `exists()` function with the capability of searching objects *recursively* --i.e. in  environments defined inside *other* environments--, as well as searching objects that are the *result of expressions*. This functionality is provided by the `obj_find()` function.  
- a simplification of the output obtained when retrieving the calling function name and the stack of calling functions, currently provided by the built-in function `sys.call()`. This functionality is provided by the `get_fun_calling()` and `get_fun_calling_chain()` functions.  
- the retrieval of the memory address of an object. This functionality is provided by the `get_obj_address()` function.  

The following sections give examples of all the above.

Please note that the package convention is to name all functions using the underscore to separate keywords (e.g. `environment_name()`, `get_obj_address()`, etc.)

## Retrieve the name of user-defined and function execution environments with `environment_name()`
In this section we illustrate the use of the `environment_name()` function.  

It extends the functionality of `environmentName()` in the base package by also retrieving the name of user-defined environments and function execution environments.
Although the name of an environment can be easily retrieved with `deparse(substitute(env1))` where `env1` is a user-defined environment, the most useful
scenario for the use of `environment_name()` is when some function tells us that an object is part of a user-defined environment, but this environment
is only given as a *memory address* (as in e.g. `<environment 0x000000000437fb40>`). In this scenario, `environment_name()` can tell us the *name* of the
environment having that memory address.  

Note that the address to name conversion also works for *function execution environments*.

The signature of the function is the following:  
`environment_name(env, envir = NULL, envmap = NULL, matchname = FALSE, ignore = NULL)`.

### Examples

#### Let's start with a few object definitions
Let's define a couple of environments and nested environments.

```{r DefineEnvironments}
env1 <- new.env()
env_of_envs <- new.env()
with(env_of_envs, env21 <- new.env())
```

So the environment `env21` is *nested* in environment `env_of_envs`.

#### Basic operation
Let's retrieve the names of the environments just defined using the `environment_name()` function. This may sound trivial because we are already typing the environment name! However, look at the output from the second call: it contains the *path* to the environment being searched for stating that `env21` is found inside environment `env_of_envs`. Below we will also see even more useful examples when we search for objects.

```{r GetEnvironmentNames, warning=FALSE}
environment_name(env1)
environment_name(env21)
```

If we already know in which environment the environment we are after is defined, we can specify it in the `envir` parameter and get their name without any path information:

***** TODO: ACTUALIZAR PACKAGE PARA QUE EN EL SEGUNDO CASO DEVUELVA "env21" Y NO
"env_of_envs$env21"! *****

```{r GetEnvironmentNamesSpecifyingLocation, warning=FALSE}
environment_name(env1, envir=globalenv())
environment_name(env21, envir=env_of_envs)
```

#### More advanced examples
Suppose now that we define another environment that points to one of the already defined environments.

```{r DefineAnotherEnvironmentThatPointsToAnExistingEnvironment}
e <- env_of_envs$env21
```

Now, let's retrieve its name:

```{r GetTheEnvironmentNameOfThisNewEnvironment}
environment_name(e)
```

What we get is the names of *all* the environments (in alphabetical order) that point to the same memory address. We can disable such behaviour of matching environments by memory address by setting the `matchname` parameter to `TRUE` so that the returned environments must match both the memory address *and* the name:

```{r CallEnvironmentNameWithMatchNameTRUE}
environment_name(e, matchname=TRUE)
```

Note that the result of this call could return more than one environment when environments sharing the same name are defined in different nvironments and they all point to the same environment. As if we defined an environment called `"e"` in environment `env_of_envs` which points to environment `e` defined in the global environment.

Finally, if we try to retrieve the environment name of a non-existing environment, we get `NULL`.

```{r CallEnvironmentNameOnNonExistingEnvironment, warning=FALSE}
environment_name(non_existing_env)
```

#### Retrieving the environment name associated with a memory address
Now suppose we have a memory address and we would like to know if that memory address represents an environment.
We can simply call `environment_name()` with the memory address as argument, as follows:

```{r ConvertMemoryAddressToEnvironmentName}
env1_obj_address = get_obj_address(env1)
environment_name(env1_obj_address)
```

Of course, in practice we would not call the `get_obj_address()` function to get the environment's memory address; we would simply type in the memory address we are after. Note that this memory address can be given either as:  

- a 16-digit address, as in: `"0000000011D7A150"`  
- a 18-digit address, as in: `"0x0000000011D7A150"`  
- either the 16-digit or 18-digit address enclosed in `< >`, as in: `"<0000000011D7A150>"` or `"<0x0000000011D7A150>"`  
- a 18-digit address preceeded by the `environment: ` keyword and enclosed in `< >`, as in: `"<environment: 0x0000000011D7A150>"`  

The latter format is particularly useful when copy & pasting the result of querying an environment object, for example when typing `env1` at the `R` command prompt, assuming that `env1` is a user-defined environment.  

If the memory address does not match any of the above formats or does not represent an environment `NULL` is returned:

```{r EnvironmentNameOfNonEnvironmentMemoryAddressIsNULL}
x = 2
environment_name(get_obj_address(x))
```

#### Retrieving the execution environment of a function
If called inside a function, `environment_name()` returns the name of the function and the environment where the function is defined.

```{r GetExecutionEnvironmentName}
with(env1,
  f <- function() {
    cat("1) We are inside function", environment_name(), "\n")
#    cat("2) We are inside function", get_fun_calling(1, showParameters=FALSE), "\n")
  }
)
env1$f()
```

## Check the existence of an object with `obj_find()`
With the `obj_find()` function we can check if an object exists in any existing environment. All environments --including system environments, packages, user-defined environments, and optionally function execution environments-- are crawled to search for the object. This includes any environments that are defined within other environments. The function returns a character array with all the environments where the object has been found.

Objects to search for can be specified either as a symbol or as a string. Ex: `obj_find(x)` and `obj_find("x")` both look for an object called "x". They can also be the result of an expression as in `v[1]`.

The function returns `NULL` if the object is not found or if the expression is invalid. For instance `obj_find(unquote(quote(x)))` returns `NULL` because the `unquote()` function does not exist in R.

The signature of the function is the following:
`obj_find(obj, envir = NULL, envmap = NULL, globalsearch = TRUE, n = 0, return_address = FALSE, include_functions = FALSE, silent = TRUE)`

#### Let's start with a few object definitions
Let's define a couple of objects in the environments already defined above:

```{r DefineObjectsInEnvironments}
x <- 5
env1$x <- 3
with(env_of_envs, env21$y <- 5)
with(env1, {
  vars_as_string <- c("x", "y", "z")
})
```

#### Basic operation
Now let's look for these objects:
```{r LookForObjects1, warning=FALSE}
cat("Object 'x' found in the following environments:\n")
obj_find(x)
cat("Object 'y' found in the following environments:\n")
obj_find(y)
```

(recall that environment `e` points to the same environment as `env_of_envs$env21`)

```{r LookForObjects2}
cat("Object 'vars_as_string' found in the following environments:\n")
obj_find(vars_as_string)
```

Let's also look for the objects defined in `vars_as_string` and `vars_quoted`.
```{r LookForObjectsWhoseNamesAreGivenInArray}
cat(paste("Object", env1$vars_as_string[1], "found in the following environments:\n"))
obj_find(env1$vars_as_string[1])  # Here we are looking for the object 'x'
cat(paste("Object", env1$vars_as_string[2], "found in the following environments:\n"))
obj_find(env1$vars_as_string[2])  # Here we are looking for the object 'y'
cat(paste("Object", env1$vars_as_string[3], "found in the following environments:\n"))
obj_find(env1$vars_as_string[3])  # Here we are looking for the object 'z' which does not exist
```

or using `sapply()`:

```{r LookForObjectsUsingSAPPLY}
cat("The objects defined in the 'env1$vars_as_string' array are found in the following environments:\n")
with(env1, sapply(vars_as_string, obj_find) )
```

Note how calling `get_obj_find()` from within the `env1` environment still searches for the objects everywhere. This is because parameter `globalsearch` is set to `TRUE`. If we set it to `FALSE` we would get:  

**** TODO: CORREGIR ESTE COMPORTAMIENTO DE get_obj_find() PORQUE EN EL CASO DE
x DEBERIA DEVOLVER 'env1'!
**** ADEMAS SI AGREGAMOS EL PARAMETRO envir=env1, el
valor de $x es "envir_orig"... WTF??

```{r LookForObjectsUsingSAPPLYNoGlobalSearch}
cat("The objects defined in the 'env1$vars_as_string' array are found in the following environments (no globalsearch):\n")
with(env1, sapply(vars_as_string, obj_find, globalsearch=FALSE) )
```

We can also search for objects given as a symbol:

```{r LookForObjectAsASymbol}
cat("Object 'x' found in the following environments:\n")
obj_find(as.name("x"))
```

Finally, we can also search for objects defined in packages:

```{r LookForObjectsDefinedInPackages}
cat("Object 'aov' found in the following environments:\n")
obj_find(aov)
```

## Retrieve the function calling chain (stack) with `get_fun_calling_chain()`
Following is an example that shows the difference between using the `sys.call()` function already available in base R and the package function `get_fun_calling_chain()` to retrieve the calling stack. 

In particular note:  
- How easy it is to check what the calling function is (just do a string comparison: `get_fun_calling() == "env1$f"`). On the contrary, when using `sys.call()` we first need to parse the output before making such a comparison. See **[this link](http://stackoverflow.com/questions/15595478/how-to-get-the-name-of-the-calling-function-inside-the-called-routine)** for more details.  
- We get a data frame containing the chain of calling functions, from the most recent call to least recent, including function parameters if desired.  

The signature of the function is the following:
`get_fun_calling_chain(n = NULL, showParameters = FALSE, silent = TRUE)`

#### Let's start with a few object definitions
1) First we define a couple of environments:

```{r DefineTwoEnvironments}
env11 <- new.env()
env12 <- new.env()
```

2) Now we define an example function `h` to be called by two different functions `f` defined in two different user-environments. This function `h` does something different depending on which function `f` was responsible for calling  it.

```{r DefineFunctionH}
with(globalenv(), 
h <- function(x, silent=TRUE) {
  fun_calling_chain = get_fun_calling_chain(silent=silent)

  # Do a different operation on input parameter x depending on the calling function
  fun_calling = get_fun_calling(showParameters=FALSE)
  if (fun_calling == "env11$f") { x = x + 1 }
  else if (fun_calling == "env12$f") { x = x + 2 }

  return(x)
}
)
```

3) Finally we define the two calling functions `f` that call `h`, respectively in environments `env11` and `env12`:

```{r DefineTwoFunctionsFInSeparateEnvironments}
with(env11,
  f <- function(x, silent=TRUE) {
    fun_calling_chain = get_fun_calling_chain()
    return(h(x, silent=silent))
  }
)

with(env12,
  f <- function(x, silent=TRUE) {
    fun_calling_chain = get_fun_calling_chain()
    return(h(x, silent=silent))
  }
)
```

### Basic operation
We now run these functions `f` and take note of their output.

- Output from `env11$f()`:

```{r RunFunctionF1, echo=FALSE}
silent = FALSE
x = 0
cat("\nWhen h(x) is called by env11$f(x=", x, ") the output is: ", env11$f(x, silent=silent), "\n", sep="")
```

- Output from `env12$f()`:

*** TODO: MODIFICAR get_fun_calling_chain() PARA QUE NO MUESTRE R_GlobalEnv$h en
el calling chain SINO SIMPLEMENTE h ***

```{r RunFunctionF2, echo=FALSE}
silent = FALSE
x = 0
cat("\nWhen h(x) is called by env12$f(x=", x, ") the output is: ", env12$f(x, silent=silent), "\n", sep="")
```

Note how easy it was (by using just a string comparison) to decide what action to take based on the `f()` function calling `h()` and perform a different operation.

## Retrieve the memory address of an object with `get_obj_address()`
Following are examples of using the `get_obj_address()` function to retrieve the memory address of an object. The object can be given either as a symbol or as an expression. In the latter case, the memory address of the result of the expression is returned. If the result is another expression, the process is not continued: the memory address of such expression is returned.

Internally this funcion first calls `obj_find()` to look for the object (using `globalsearch=TRUE`) and then retrieves the object's memory address, if the object is found, showing the name of all the environments where the object was found. If the object is not found, `NULL` is returned.  

The signature of the function is the following:
`get_obj_address(obj, envir = NULL, envmap = NULL, n = 0, include_functions = FALSE)`

### Examples
The following two calls return the same result:

```{r GetObjectAddress1}
cat("Output of 'get_obj_address(x)':\n")
get_obj_address(x)
cat("\nOutput of 'with(env1, get_obj_address(x))':\n")
with(env1, get_obj_address(x))
```

Note especially the last case, where calling `get_obj_address()` from within the `env1` environment still searches for the object everywhere.

We can restrict the memory address returned by making the environment where the object is located explicit --by either using the `$` notation or the `envir` parameter of `get_obj_address()`. In this case only the address of the specified object is returned, even if other objects with the same name exist within the specified environment. A few examples follow:

```{r GetObjectAddress2}
get_obj_address(env1$x)
get_obj_address(x, envir=env1)
with(env1, get_obj_address(x, envir=env1))
```

Note there is a slight difference between calling `get_obj_address()` using the `$` notation and calling it with the `envir=` parameter: in the latter case, the result is an _unnamed_ array.    

&nbsp;
&nbsp;

Suppose now the object is an expression referencing three potential existing objects as strings, more specifically an array:

```{r GetNonExistentObjectAddress}
vars = c("x", "y", "nonexistent")
get_obj_address(vars[1], envir=env1)
sapply(vars, get_obj_address)
```

(recall that environment `e` points to the same environment as `env_of_envs$env21`)

We can check that the memory address is correct by running the internal function `envnames:::address()` which calls a C function that retrieves the memory address of a variable:

```{r CheckMemoryAddressIsCorrect}
envnames:::address(env1$x)
envnames:::address(e$y)
```

WARNING: If we call `envnames:::address()` on a non-existent object inside an environment to which we refer using the `$` notation, we still get a memory address and this is the memory address of `NULL`. In fact the following calls give the same result:

```{r MemoryAddressOfNULL}
envnames:::address(env1$nonexistent)
envnames:::address(NULL)
```
However, this is not the case if we use the `get_obj_address()` function as it returns `NULL` when the object is not found:

```{r MemoryAddressOfNonExistentIsNULL}
get_obj_address(env1$nonexistent)
```
